Influence of journal data archiving mandates on author attitudes, experiences,

### 2018-03-22 manuscript.Rmd
methods, results, and discussion updates

## The promise of data sharing

@@

## Data archiving journal policies

@@

## Analysis

The responses of authors who published in JDAP-adopting journals were compared, over time, to  the responses of authors who published in similar journals that did not adopt JDAP.

The primary analysis will be a seven-point Likert scale asking how strongly authors believe:  “It is the community norm in my field to publicly share datasets upon study publication by archiving all datasets online, for use by anyone for any purpose.”  About twelve other survey questions will allow secondary and exploratory analyses.

@@

The following plots show the responses regarding the attitudes towards data sharing and archiving for participants who (i) published in journals which have note adopted the JDAP so far ('No JDAP'), (ii) published in journals which had not _yet_ adopted the JDAP before publication ('Before JDAP'), and (iii) journals which had adopted the JDAP before publication ('After JDAP).

@@

The potential effect of JDAP adoption on authors' attitudes is might be confounded by two other factors. Firstly, groups of researchers publishing in journals which are not planning to adopt the JDAP (or at least have not done so yet) may inherently differ from groups publishing in journals which were open to adopting the JDAP. Secondly, attitudes regarding data archiving may improve over time.

[TODO: REPHASE PARAGRAPHS BELOW]

In order to test for an effect of policy change on attitudes while controlling for potential confounding effects, we used an ordinal logistic regression model with three predictors: (i) journal type _(cNonJDAPJournal)_, (ii) (centered) publication date in years _(cDate)_, and (iii) a variable indicating whether the article was published after the the journal adopted a data archiving policy _(cAfterPolicyChange)_.

@@

m4 <- polr(respQ1 ~ 1 + cAfterPolicyChange + cNonJDAPJournal + cDate, data=d , Hess=TRUE)
# summary(m4)
m4.justchange <- polr(respQ1 ~ 1 + cAfterPolicyChange, data=d , Hess=TRUE)
m4.nochange <- polr(respQ1 ~ 1 + cNonJDAPJournal + cDate, data=d , Hess=TRUE)
suppressMessages( ci4 <- confint(m4) )
table4 <- exp(cbind(OR = coef(m4), ci4)) %>% cbind(., t=coef(summary(m4))[1:nrow(.),'t value'])
#table4 <- cbind(lOR = coef(m4), ci4) %>% cbind(., t=coef(summary(m4))[1:nrow(.),'t value'])

m5 <- polr(respQ2 ~ 1  + cAfterPolicyChange + cNonJDAPJournal + cDate, data=d, Hess=TRUE)
m5.justchange <- polr(respQ2 ~ 1 + cAfterPolicyChange , data=d, Hess=TRUE)
m5.nochange <- polr(respQ2 ~ 1 + cNonJDAPJournal + cDate, data=d, Hess=TRUE)
# summary(m5)
suppressMessages( ci5 <- confint(m5) )
table5 <- exp(cbind(OR = coef(m5), ci5)) %>% cbind(., t=coef(summary(m5))[1:nrow(.),'t value'])

aov.m4.justchange <- anova(m4, m4.justchange)
aov.m4.nochange <- anova(m4, m4.nochange)
aov.m5.justchange <- anova(m5, m5.justchange)
aov.m5.nochange <- anova(m5, m5.nochange)

@@

The following table shows the estimated coefficients and 95\%-confidence intervals (transformed to odds ratios) for responses to *question 1*. Effects are statistically significant at the $\alpha=0.05$ level when the transformed confidence intervals do not contain $1$. Confidence intervals were obtained by profile likelihood. According to the model estimates, the introduction of the JDAP data archiving policy significantly increases the odds of a positive response regarding *data sharing* by a factor of $1.36$ (CI=$[1.16; 1.60]$), while the remaining two predictors do not show any statistically significant effects.

@@

We conducted two likelihood-ratio tests to assess the effect of change in policy and the other predictors on reported attitudes. The first likelihood-ratio test, comparing the full three-predictor model with a reduced model _without_ the effect of policy change revealed a significantly better fit for the full model ($\chi^2(`r aov.m4.nochange[["   Df"]][2]`)=`r aov.m4.nochange[["LR stat."]][2]`$, $p<0.001$). A second likelihood-ratio test comparing the full model with a reduced model using _only_ the policy change predictor (_cAfterPolicyChange_) revealed no statistically significantly difference ($\chi^2(`r aov.m4.justchange[["   Df"]][2]`)=`r aov.m4.justchange[["LR stat."]][2]`$, $p=`r aov.m4.justchange[["Pr(Chi)"]][2]`$).

@@

Similarily, the following table shows the estimated coefficients and 95\%-confidence intervals (transformed to odds ratios) for responses to *question 2*. According to the model estimates, the introduction of the JDAP data archiving policy significantly increases the odds of a positive response regarding *data archiving* by a factor of $1.57$ (CI=$[1.33; 1.90]$), while the remaining two predictors do not show any statistically significant effects.

Here too, we conducted two likelihood-ratio tests to assess the effect of change in policy and the other predictors on reported attitudes. The first likelihood-ratio test, comparing the full three-predictor model with a reduced model _without_ the effect of policy change revealed a significantly better fit for the full model ($\chi^2(`r aov.m5.nochange[["   Df"]][2]`)=`r aov.m5.nochange[["LR stat."]][2]`$, $p<0.001$). A second likelihood-ratio test comparing the full model with a reduced model using _only_ the policy change predictor (_cAfterPolicyChange_) revealed no statistically significantly difference ($\chi^2(`r aov.m5.justchange[["   Df"]][2]`)=`r aov.m5.justchange[["LR stat."]][2]`$, $p=`r aov.m5.justchange[["Pr(Chi)"]][2]`$).

@@

data policies are becoming more prevalent: seven journals announced data archiving mandates during the course of the survey (and another three journals implemented policies shortly after the suvey completed).

@@

Even when there are journal policies, often not sufficient to achieve reproducability for the majority of studies `r citep(c(bib["roche2015public"], bib["stodden2018an-empi"], bib["naudet2018data-sh"]))`

suggesting author personality `r citep(c(bib["linek2017data-sh"]))` or social norms `r citep(c(bib["anagnostou2015when-da"], bib["womack2015researc"]))` may play a larger factor than journal policy.

Findings here are consistent with other findings that authors more likely to share data if they had prior experience sharing or reusing data
`r citep(c(bib["Piwowar_2011"]))`.

@@

Similarily, although of course data archiving has a cost, it also has a good return on investment in terms of research output `r citep(c(bib["piwowar2011data-ar"]))`.

@@

cite this? `r citep(c(bib["van-tuyl2016water-w"]))`. 
also maybe this?  `r citep(c(bib["fetterman2015the-rep"]))`.

still pulling these in to discussion notes `r citep(c(bib["budin-ljosne2014data-sh"],bib["lin2014recomme"],bib["roche2014trouble"],bib["vines2013mandate"],bib["pearce2011data-sh"],bib["alsheikh-ali2011public"]))`

@@

- IRB (now in dropbox)
- Recruitment emails (now in dropbox)
- Journal policies (instructions to authors as of January 29, 2011): zip of html (now in dropbox)
- Survey questions (now in dropbox)
- [Survey responses](https://github.com/hpiwowar/jdap/blob/master/data/raw_data/JDAP_Survey_Data.csv)
- [Full text comments](https://github.com/hpiwowar/jdap/blob/master/data/full_text_comments.txt)
- [Contact corresponding authors scripts](https://github.com/hpiwowar/contact_corresponding)
- [Manuscript R files](https://github.com/hpiwowar/jdap)

>>>>

## Data archiving policies

@@

@@

## Analysis approach

The primary analysis will be based on responses to be a seven-point Likert scale asking how strongly authors believe:  “It is the community norm in my field to publicly share datasets upon study publication by archiving all datasets online, for use by anyone for any purpose.”  Our goal is to see if author's responses to this question were influenced by publishing in a journal that had implemented JDAP.

The potential effect of JDAP adoption on authors' attitudes is might be confounded by two other factors. First, groups of researchers publishing in journals which are not planning to adopt JDAP may inherently differ from groups publishing in journals which were open to adopting the JDAP. Second, attitudes regarding data archiving may improve over time for all authors, independent of journal or policy.

In order to test the effect of policy change on attitudes while controlling for potential confounding effects, we used an ordinal logistic regression model with three predictors: (i) whether the jounal adopted JDAP _(cNonJDAPJournal)_, (ii) the date the authors published their paper _(cDate)_, and (iii) a variable indicating whether the article was published after the the journal adopted a data archiving policy _(cAfterPolicyChange)_.  Confidence intervals were obtained by profile likelihood.  If the confidence intervals do not contain $1$ for the _(cAfterPolicyChange)_ variable then the policy would have had an affect on attitudes, independent of publication date or whether the journal was the type to adopt a JDAP policy.

We further tested the policy correlation by comparing the fit of two models, one model with all three of the above variables, and one model with two variables excluding the policy change predictor (_cAfterPolicyChange_).  If, using a likelihood-ratio test, the full model was a statistically better fit than the model without the policy change variable, it would support the hypothesis that the differences in authors attitudes were related to the policy adoption.

Twelve other survey questions allowed secondary and exploratory analyses.

@@

The following plots show the responses regarding the attitudes towards data sharing and archiving for participants who (i) published in journals which have note adopted the JDAP so far ('No JDAP'), (ii) published in journals which had not _yet_ adopted the JDAP before publication ('Before JDAP'), and (iii) journals which had adopted the JDAP after publication ('After JDAP).

@@

The following plots show the responses regarding the attitudes towards data sharing and archiving for participants who (i) published in journals which have note adopted the JDAP so far ('No JDAP'), (ii) published in journals which had not _yet_ adopted the JDAP before publication ('Before JDAP'), and (iii) journals which had adopted the JDAP after publication ('After JDAP).

According to the model estimates, the introduction of the JDAP data archiving policy significantly increases the odds of a positive response regarding *data sharing* by a factor of $1.36$ (CI=$[1.16; 1.60]$), while publishing date and whether the journal ever adopted a JDAP policy did not show any statistically significant effects.  The estimated coefficients and confidence intervals are shown below.

@@

m4 <- polr(respQ1 ~ 1 + cAfterPolicyChange + cNonJDAPJournal + cDate, data=d , Hess=TRUE)
# summary(m4)
m4.nochange <- polr(respQ1 ~ 1 + cNonJDAPJournal + cDate, data=d , Hess=TRUE)
suppressMessages( ci4 <- confint(m4) )
table4 <- exp(cbind(OR = coef(m4), ci4)) %>% cbind(., t=coef(summary(m4))[1:nrow(.),'t value'])
#table4 <- cbind(lOR = coef(m4), ci4) %>% cbind(., t=coef(summary(m4))[1:nrow(.),'t value'])

m5 <- polr(respQ2 ~ 1  + cAfterPolicyChange + cNonJDAPJournal + cDate, data=d, Hess=TRUE)
m5.nochange <- polr(respQ2 ~ 1 + cNonJDAPJournal + cDate, data=d, Hess=TRUE)
# summary(m5)
suppressMessages( ci5 <- confint(m5) )
table5 <- exp(cbind(OR = coef(m5), ci5)) %>% cbind(., t=coef(summary(m5))[1:nrow(.),'t value'])

aov.m4.nochange <- anova(m4, m4.nochange)
aov.m5.nochange <- anova(m5, m5.nochange)

@@

@@

The result of two likelihood-ratio tests supported these findings.  The first likelihood-ratio test, comparing the full three-predictor model to a reduced model _without_ the effect of policy change revealed a significantly better fit for the full model ($\chi^2(`r aov.m4.nochange[["   Df"]][2]`)=`r aov.m4.nochange[["LR stat."]][2]`$, $p<0.001$), suggesting that the date of publication and whether a journal adopted JDAP are not as important in to an author's opinion about *data sharing* the data as whether the author published after a JDAP policy was introduced.

@@

We found the similar results when we looked at the author's responses to community norms about *data archiving* rather than data sharing. The following table shows the estimated coefficients and 95\%-confidence intervals (transformed to odds ratios). According to the model estimates, the introduction of the JDAP data archiving policy significantly increases the odds of a positive response regarding *data archiving* by a factor of $1.57$ (CI=$[1.33; 1.90]$).  Being a nonJDAP journal is a significant predictor in this case, though with a much lower odds ratio.

Comparing the full three-predictor model with a reduced model _without_ the effect of policy change revealed a significantly better fit for the full model ($\chi^2(`r aov.m5.nochange[["   Df"]][2]`)=`r aov.m5.nochange[["LR stat."]][2]`$, $p<0.001$), indicating again that the policy change was a statistically significant predictor of the author's responses about data archiving attitudes.

@@

We found that more authors believed that data sharing and data archiving were norms in their field after they published in a journal with a JDAP policy.

Data archiving policies are becoming more prevalent: seven journals announced data archiving mandates during the course of the survey (and another three journals implemented policies shortly after the suvey completed).

@@

Journal policies do help  `r citep(c(bib["vines2013mandate"], bib["Piwowar_2011"], bib["alsheikh-ali2011public"]))`, though often not sufficient to achieve data availability for the majority of studies `r citep(c(bib["Piwowar_2011"], bib["alsheikh-ali2011public"], bib["roche2015public"], bib["stodden2018an-empi"], bib["naudet2018data-sh"]))`.

This suggests author personality `r citep(c(bib["linek2017data-sh"]))` or social norms `r citep(c(bib["anagnostou2015when-da"], bib["womack2015researc"]))` may play a larger factor.

@@

Similarily, the free text comments suggested authors are concerned about the cost of data archiving.  It may help to spread the word that money research funders spend on data archiving is good return on investment in terms of research output `r citep(c(bib["piwowar2011data-ar"]))`.

@@

Our findings are consistent with previous studies that found authors were more likely to share data if they had prior experience sharing or reusing data: `r citep(c(bib["Piwowar_2011"]))`.

Several recommendations have been made for journals seeking to form a data archiving policy `r citep(c(
bib["doi:10.1002/asi.23336"],
bib["van-tuyl2016water-w"],
bib["roche2014trouble"],
bib["lin2014recomme"]))`

@@

- IRB (now in dropbox)
- Recruitment emails (now in dropbox)
- Journal policies (instructions to authors as of January 29, 2011): zip of html (now in dropbox)
- Survey questions (now in dropbox)
- [Survey responses](https://github.com/hpiwowar/jdap/blob/master/data/raw_data/JDAP_Survey_Data.csv)
- [Full text comments](https://github.com/hpiwowar/jdap/blob/master/data/full_text_comments.txt)
- [Contact corresponding authors scripts](https://github.com/hpiwowar/contact_corresponding)
- [Manuscript R files](https://github.com/hpiwowar/jdap)
- [News blog about the study](https://studyonimpactofjournaldatapolicies.wordpress.com/)
